---
title: "EDA and Visuals"
author: "Aleksander Rodriguez"
date: "2024-04-24"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exploratory Data Analysis
Within this exploratory analysis statistical tests and visuals will be created based
on the data provided. Variables will be labeled to be strong or weak depending on 
their p-values, f-values, and correlation numbers. The visuals will be made based
off some of the strongest variables to give information and the relation the that 
variable has to BMI.


```{r}
# EDA Part 

obesity <- read.csv("C:/Users/Alek4/OneDrive/Documents/CsvFiles/ObesityDataSet.csv")

# Wanted to see how the data is spread out. I didn't want the data to be lopsided 
# toward one demographic which could make models less accurate. The table 
# shows that there is a good spread within the data.
table(obesity$Obese_level)

# Adding the BMI column by calculating it with weight and height. 
# BMI stands for Body Mass Index. It measures an individual's body
#weight relative to their height. 
obesity$BMI <- obesity$Weight/(obesity$Height ^ 2)
obesity$Weight_Status <- ifelse(obesity$BMI  >= 30, 1,0)
head(obesity)
```

## Testing Variables

Checking if the variable's represent a good p-value and f-value
so we can make good analysts over the data given. Correlation test's will also 
be used for continuous data to determine if the data is good to use. A strong 
relation or correlation is what we are looking for.

Age has strong a correlation 
```{r}
cor_results <- cor.test(obesity$BMI, obesity$Age)
cor_results
```
Not a strong relation between BMI and gender
```{r}
t.test(BMI ~ Gender, data = obesity)
```
Weight has a strong correlation to BMI
```{r}
cor_results <- cor.test(obesity$BMI, obesity$Weight)
cor_results
```
Height has a strong correlation but is not as strong as others to BMI
```{r}
cor_results <- cor.test(obesity$BMI, obesity$Height)
cor_results

```
Alcohol has a strong relation with BMI
```{r}
anova_results <- aov(BMI ~ factor(Alcohol), data = obesity)
summary(anova_results)
```
High_Cal has a strong relation with BMI
```{r}
t.test(BMI ~ High_Cal, data = obesity)
```
Veggies_consumed has a strong correlation to BMI
```{r}
cor.test(obesity$BMI, obesity$Veggies_consumed)
```

Meals_Daily surprisingly does not have a good correlation with BMI
```{r}
cor.test(obesity$BMI, obesity$Meals_Daily)
```
Monitor_Cals has a strong relation with BMI
```{r}
t.test(BMI ~ Monitor_Cals, data = obesity)
```
Smoke has a weak relation with BMI
```{r}
t.test(BMI ~ SMOKE, data = obesity)
```
Water_consumed has a strong correlation but is not as strong as others
```{r}
cor.test(obesity$BMI, obesity$Water_Consumed)
```
family_history_with_overweight has a strong relation with BMI
```{r}
t.test(BMI ~ family_history_with_overweight, data = obesity)
```
Physical_Activity has a strong correlation to BMI
```{r}
cor.test(obesity$BMI, obesity$Physical_Activity)
```
TUE has a strong correlation but is not as strong as others
```{r}
cor.test(obesity$BMI, obesity$TUE)
```
Transportation_Used has a strong relation to BMI 
```{r}
anova_results <- aov(BMI ~ factor(Tranportaion_Used), data = obesity)
summary(anova_results)
```
## Summary of Testing

  Throughout the EDA of this data, the discovery of many variables with a strong 
correlation and relation to BMI were found. The EDA process found that the best 
variables are the amount of consumption of alcohol, veggies, and high-calorie meals. 
Others were the amount of physical activity, family history of being overweight, and 
the mode of transportation used. These variables make sense due to the fact these
correlate with healthy habits. Eating veggies and low calorie meals, with a significant
amount of physical activity can help with weight loss. Mode of transportation can also
help lower BMI levels due having to walking or biking places. Having family history of being
overweight is another variable that makes sense. If one grows up in a house with others
that do not maintain a low BMI then most likely the eating habits of their family 
members would ware off on them. Drinking water was a strong correlation but was not 
as important as others.

  A surprising variable, that was one of the worst in finding a correlation with BMI,
was the amount of meals one eats daily. This could provide insight into the fact that 
it might not be about how often one eats, but rather what one eats. For example, if 
you eat five times a day but consume veggies and low-calorie food, you might not 
increase your BMI, compared to eating three times a day with high-calorie food. 
Smoking and usage of technology were other weak variables.

  Most of the variables made sense and had a reasonable relation to BMI. With the
conclusion of the EDA process, the process of building visuals and models based 
on the strong variables can commence to help communicate findings within the data.Then
eventually give ideas, actions, and reasons of how to lower their BMI according to the data.


## Visuals to Illustrate The EDA

The following graph represents the importance of physical activity. The y-axis contains
ones physical activity, x-axis contains ones BMI, and the dots calories according to 
their obesity level. It shows show that to lower ones BMI and stray away form obesity 
one should be more physically active. The graph gives evidence to why someone should 
keep active physically. 
```{r}
library(ggplot2)

ggplot(obesity, aes(x = BMI, y = Physical_Activity, color = factor(Obese_level))) +
  geom_point() + 
  labs(x = "BMI", y = "Physical Activity", 
       title = "Relationship Between BMI and Physical Activity by Obesity Level") +
  scale_color_discrete(name = "Obesity Level")
```



The following graph shows that higher calories within one's meals can effect their BMI's 
by increasing it. Higher calories results in a higher BMI. This can also give 
evidence as to why the amount of meals one eat does not matter but what they eat
does. Less calories in one's meals is better for ones BMI. Food companies should be 
encouraged to push lower calorie meals and the public should be informed on how much
calorie intake is good for their body.
```{r}
ggplot(obesity, aes(x = High_Cal, y = BMI, fill = High_Cal)) +
  geom_boxplot() +
  labs(x = "High Caloric Intake", y = "BMI", 
       title = "Relationship Between High Caloric Intake and BMI") +
  theme_minimal()
```



The following graph shows the relation between family history of being overweight
and ones BMI. The findings illustrate the strong correlation between these two variables.
It gives the idea that if you were born into a family that is overweight you 
will have a higher chance to be at a high BMI. If your family is overweight and maintains
an unhealthy lifestyle you might partake or pick up those unhealthy habits as well. 
```{r}
ggplot(obesity, aes(x = family_history_with_overweight, y = BMI, fill = 
                      family_history_with_overweight)) +
  geom_boxplot() + 
  labs(x = "Family History Of Being Overweight", y = "BMI", title = "BMI vs. Age with Obesity Level") +
  scale_color_discrete(name = "Obesity Level")
```




